---
title: "Pressure or Overconfidence: Why do batters get out?"
author: "Dr Preet Deep Singh"
affiliation: "Invest India"
date: "12/01/2022"
output: 
  bookdown::pdf_document2:
    toc: no
    citation_package: natbib
    number_sections: yes
    keep_tex: yes
abstract: "Batters would be overconfident after a boundary and under pressure after a dot ball. The pressure would be higher in the second innings and overconfidence might dominate in the first innings. We look at 1,50,460 deliveries and find that batters are more likely to get out after a dot ball in the second innings as compared to the first innings. Similarly, batters are more likely to be dismissed after a boundary in the first innings than in the second innings. "
documentclass: "article"
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
bibliography: references.bib
link-citations: true
csl: apa.csl
---

```{r setup, echo=FALSE, message=FALSE,warning=FALSE, tidy=TRUE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(dplyr)
library(stargazer)
library(bookdown)
library(ARTofR)
library(ggplot2)
#library(BSDA) #For z test
#library(dgof) #For z test
library(readr)
df1 <- read_csv("deliveries.csv")
df1<-cbind(1:length(df1$match_id),df1)
colnames(df1)[1]<-"sno"


## Subset of wickets
dfw<-subset(df1,!is.na(player_dismissed)) #To get all previous balls
dfw<-subset(dfw, over+ball >2) #no first ball wickets here
#df2<-data.frame()
#for(i in 1:length(dfw$sno)){
  
#  df2[i,1:22]<-df1[dfw$sno[i]-1,1:22]
#}
#write.csv(df2,"dfbeforewicket.csv")
df2 <- read_csv("dfbeforewicket.csv")
df2<-df2[,2:23]# Somehow we had a bad column in the beginning. This takes care of that.
df3<-rbind(dfw,df2)
df3<-df3[order(df3$sno,decreasing=FALSE),]
```

\section{Introduction}
A batter could feel more confident after a boundary and may hit an improper shot allowing bowlers to take that wicket. Another reason for an improper shot could be pressure. A dot ball causes pressure. It could also indicate good bowling and/or tight fielding. In order to isolate the role of pressure we look at the difference between innings. A dot ball is likely to cause more pressure in the second innings while chasing rather than in the first innings while scoring. Similarly one is likley to be over-confident in the first innings while setting the score than in the second innings while chasing a score.

\section{Data}

Data source: https://www.kaggle.com/ash316/let-s-play-cricket/data

```{r datasummary, results='asis'}

stargazer(df1[2:20], title="This table shows the summary of all the deliveries used for the study",header=FALSE)
```
We get `r length(df1$sno)` deliveries from `r length(unique(df1$match_id))` matches. These span all forms of cricket: national, league, international, one days and tests. This data is sourced from Kaggle\footnote{https://www.kaggle.com/ash316/let-s-play-cricket/data}. I identify the deliveries that have a wicket. I create a subset of the data that only contains deliveries that had wickets. I then compile a subset of the data that only includes balls immediately preceding the wicket ball. I then look at the frequency distribution of the scores in that dataset. Since the number of deliveries is not the same, the graphs are not comparable. I therefore look at the percentage break up of the composition of scores. There are eight values possible: 7 runs and wicket. In case of a no-ball or wide, the runs column indicates only the extra number of runs. 

I run another version with the previous ball faced by the same batter. For example, if a batter was bowled on the first ball of the second over, then I either ignore this entry or look at only those cases where the last ball indicated 1/3 runs scored by the batter. This allows me to isolate pressure at the player level which might be higher than that at the team level. 


\section{Results}
\subsection{Innings wise}


```{r first }
df4<-df2 #For safety
#xxx_box1("For Same Batter")

#...............................................................................
#                                                                              .
#  For Same Batter                                                            .
#                                                                              .
#...............................................................................

# df2a<-subset(df2, batter_runs!=1) #To remove singles
# df2b<-subset(df2, ball==1) #To remove cases where there was a strike change. So previous ball was single and over change or new over.
# Strike change = 1,3,5 which si %% 2 != 0
df2d<-subset(df2, batsman_runs%%2==0) # Only cases where strike has not changed 

df2c<-subset(df2,ball==6) #Only previous over , so first ball wickets
df2c2<-subset(df2c, batsman_runs%%2!=0 )

df2e<-rbind(df2d,df2c2) # This is our file
df2<-df2e #more for nomenclature
df2$total_runsOrg<-df2$total_runs
df2$total_runsOrg<-as.numeric(as.character(df2$total_runsOrg))
df2$total_runs<-ifelse(is.na(df2$player_dismissed),df2$total_runs,"W")


#...............................................................................
#                                                                              .
#  Coming back to DF2                                                           .
#                                                                              .
#...............................................................................


prevballsummary<-df2%>%group_by(total_runs,inning)%>% tally() #To get a count of all things that happened prevoiously
prevballsummary1<-filter(prevballsummary,inning==1)
prevballsummary1$perc<-prevballsummary1$n/sum(prevballsummary1$n)
prevballsummary2<-filter(prevballsummary,inning==2)
prevballsummary2$perc<-prevballsummary2$n/sum(prevballsummary2$n)
prevballsummary<-select(prevballsummary,total_runs,n)
```

In the graphs we can see that in 42% of the cases the wicket is preceded by a dot ball in the first innings whereas it is seen in 45% of the cases in the second innings. Along with 7% of the balls that had a wicket on the previous ball. Just these two account for more than half the total wickets and that indicates that pressure has a serious role to play in dismissals. 
```{r firstplots}

#plot1<-ggplot(data=prevballsummary,aes( x=reorder(total_runs,-n),y=n))+geom_bar(stat="identity") # Somehow this is not in order
plot1<-ggplot(data=prevballsummary,aes( x=total_runs,y=n))+geom_bar(stat="identity") +ggtitle("Frequency distribution of outcomes on the ball preceding the wicket")
#plot1
#plot2<-ggplot(data=prevballsummary1,aes( x=reorder(total_runs,-n),y=n))+geom_bar(stat="identity")+ggtitle("Frequency distribution of outcomes on the ball preceding the wicket in first innings") # Only singles are extras
plot2b<-ggplot(data=prevballsummary1,aes( x=reorder(total_runs,-n),y=perc))+geom_bar(stat="identity")+ggtitle("Frequency distribution of outcomes on the ball preceding \n the wicket in the first innings") + xlab("Outcome")+ylab("Percentage")+
  geom_text(aes(label=round(perc,2)), position=position_dodge(width=0.9), vjust=-0.25)# Only singles are extras
plot2b
#plot3<-ggplot(data=prevballsummary2,aes( x=reorder(total_runs,-n),y=n))+geom_bar(stat="identity")+ggtitle("Before wicket in second innings") # Only singles are extras
plot3b<-ggplot(data=prevballsummary2,aes( x=reorder(total_runs,-n),y=perc))+geom_bar(stat="identity")+ggtitle("Frequency distribution of outcomes on the ball preceding \n the wicket in the second innings") +xlab("Outcome")+ylab("Percentage")+
  geom_text(aes(label=round(perc,2)), position=position_dodge(width=0.9), vjust=-0.25)# Only singles are extras
plot3b

```

```{r fifthdifferences}

df2$Boundary<-ifelse(df2$total_runs==4,1,ifelse(df2$total_runs==6,1,0))
df2$Four<-ifelse(df2$total_runs==4,1,0)
df2$Six<-ifelse(df2$total_runs==6,1,0)

df1$Boundary<-ifelse(df1$total_runs==4,1,ifelse(df1$total_runs==6,1,0))
df1$Four<-ifelse(df1$total_runs==4,1,0)
df1$Six<-ifelse(df1$total_runs==6,1,0)


df2$Zero<-ifelse(df2$total_runs==0,1,0)
df2second<-subset(df2,inning==2)
df2first<-subset(df2,inning==1)
df1$Zero<-ifelse(df1$total_runs==0,1,0)
df1second<-subset(df1,inning==2)
df1first<-subset(df1,inning==1)



```

I examine the difference in the two innings with a closer lens.
We do a T test to examine the difference in the runs scored by the batter on the ball before the wicket in the first as well as the second innings. We find the mean value of runs scored by the batter on the ball preceeding the wicket in the first innings as `r round(t.test(df2first$batsman_runs,df2second$batsman_runs)$estimate[1],2)` and `r round(t.test(df2first$batsman_runs,df2second$batsman_runs)$estimate[2],2)` in the second innings. The p value of the difference is `r t.test(df2first$batsman_runs,df2second$batsman_runs)$p.value` making it statistically very significant. Similarly we code a column as 1/0. 1 when the score on the ball is zero and 0 in all other cases. We find that the mean value in the first innings, as per the ball preceeding the wicket is `r round(t.test(df2first$Zero,df2second$Zero)$estimate[1],2)` and in the second innings it is `r round(t.test(df2first$Zero,df2second$Zero)$estimate[2],2)` with a p value of `r t.test(df2first$Zero,df2second$Zero)$p.value` making it statistically very significant.


I do a further test to examine whether the balls preceeding the wicket are similar to the rest of the balls in the match. 


```{r secondmatch}
df1$total_runs<-ifelse(is.na(df1$player_dismissed),df1$total_runs,"W")
prevballsummaryall<-df1%>%group_by(total_runs,inning)%>% tally()
prevballsummaryall1<-filter(prevballsummaryall,inning==1)
prevballsummaryall1$perc<-prevballsummaryall1$n/sum(prevballsummaryall1$n)
prevballsummaryall2<-filter(prevballsummaryall,inning==2)
prevballsummaryall2$perc<-prevballsummaryall2$n/sum(prevballsummaryall2$n)
#plot4<-ggplot(data=prevballsummaryall,aes( x=reorder(total_runs,-n),y=n))+geom_bar(stat="identity") +ggtitle("Throughout the match") # Somehow this is not in order
#plot4<-ggplot(data=prevballsummaryall,aes( x=total_runs,y=n))+geom_bar(stat="identity")  + ggtitle("Throughout the match") 
#plot4
plot5<-ggplot(data=prevballsummaryall1,aes( x=total_runs,y=perc))+geom_bar(stat="identity")  + ggtitle("Frequency distribution of outcomes on all balls \n in the first innings")+xlab("Outcome")+ylab("Percentage")+
  geom_text(aes(label=round(perc,2)), position=position_dodge(width=0.9), vjust=-0.25)# Only singles are extras
plot5
plot6<-ggplot(data=prevballsummaryall2,aes( x=total_runs,y=perc))+geom_bar(stat="identity")  + ggtitle("Frequency distribution of outcomes on all balls \n in the second innings")+xlab("Outcome")+ylab("Percentage")+
  geom_text(aes(label=round(perc,2)), position=position_dodge(width=0.9), vjust=-0.25)# Only singles are extras
plot6
# `r round(mean(subset(df1,inning==1)$total_runs,na.rm=TRUE),2)` 
#is `r round(mean(df2first$total_runsOrg,na.rm=TRUE),2)` with a 
# p value of `r round(t.test(subset(df1,inning==1)$total_runs,df2first$total_runsOrg)$p.value,5)`. Similarly  ##SOmething here.
# `r round(t.test(df2first$batsman_runs,df2second$batsman_runs)$estimate[1],2)` and `r round(t.test(df2first$batsman_runs,df2second$batsman_runs)$estimate[1],2)` in the 
```

In these graphs I show that the balls preceding the wicket are different from the overall composition of the innings. For example if the entire match is composed of deliveries with outcomes similar to those on deliveries preceding the wicket then there would be very little predictive value. The graphs show a significantly different composition of outcomes. Statistically the mean value of the outcome on all balls of the first innings is `r round(mean(df1first$total_runs,na.rm=TRUE),2)`
and that for all balls preceding wickets in the first innings is `r round(mean(df2first$total_runsOrg,na.rm=TRUE),2)` with a significant p value.

\subsection{Looking at Zeros}
The mean value of the Zero column for all the deliveries in the first innings is `r round(mean(df1first$Zero,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the first innings is `r round(mean(df2first$Zero,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1first$Zero,df2first$Zero)$p.value,2)`
The mean value of the Zero column for all the deliveries in the second innings is `r round(mean(df1second$Zero,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the second innings is `r round(mean(df2second$Zero,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1second$Zero,df2second$Zero)$p.value,5)`. Note that a higher mean value here means more zeroes because each row with zero score is represented as 1. It shows a substantially higher number of dot balls before a wicket as compared to the rest of the match. 
\subsection{Looking at Boundaries}
The mean value of the Boundary column for all the deliveries in the first innings is `r round(mean(df1first$Boundary,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the first innings is `r round(mean(df2first$Boundary,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1first$Boundary,df2first$Boundary)$p.value,2)`
The mean value of the Boundary column for all the deliveries in the second innings is `r round(mean(df1second$Boundary,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the second innings is `r round(mean(df2second$Boundary,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1second$Boundary,df2second$Boundary)$p.value,5)`.
It is clear that while the total mean boundaries in the entire innings is similar for the first and second innings, the number of boundaries preceding a wicket is much higher in the first innings and this difference is statistically significant. The p value for the difference between the two innings for boundaries preceding a wicket is `r round(t.test(df2first$Boundary,df2second$Boundary)$p.value,5)`

\subsubsection{Looking at Fours}
The mean value of the Four column for all the deliveries in the first innings is `r round(mean(df1first$Four,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the first innings is `r round(mean(df2first$Four,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1first$Four,df2first$Four)$p.value,2)`
The mean value of the Four column for all the deliveries in the second innings is `r round(mean(df1second$Four,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the second innings is `r round(mean(df2second$Four,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1second$Four,df2second$Four)$p.value,5)`.
Similar to the above while the mean fours in the entire innings is similar for the first and second innings, the number of fours preceding a wicket is much higher in the first innings and this difference is statistically significant(between innings and balls before wicket). The p value for the difference between the two innings for fours preceding a wicket is `r round(t.test(df2first$Four,df2second$Four)$p.value,5)` between fours preceding wickets in the first innings versus second innings.

\subsubsection{Looking at Sixes}
The mean value of the Six column for all the deliveries in the first innings is `r round(mean(df1first$Six,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the first innings is `r round(mean(df2first$Six,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1first$Six,df2first$Six)$p.value,2)`
The mean value of the Six column for all the deliveries in the second innings is `r round(mean(df1second$Six,na.rm=TRUE),2)` versus that of all deliveries preceding a wicket in the second innings is `r round(mean(df2second$Six,na.rm=TRUE),2)` with a p value of 
`r round(t.test(df1second$Six,df2second$Six)$p.value,5)`.
Similar to the above while the mean sixers in the entire innings is similar for the first and second innings, the number of sixers preceding a wicket is much higher in the first innings and this difference is statistically significant(between innings and balls before wicket). The p value for the difference between the two innings for sixers preceding a wicket is `r round(t.test(df2first$Six,df2second$Six)$p.value,5)` between sixers preceding wickets in the first innings versus second innings.







\section{Conclusion and Limitations}
We show that batters are more likely to be under pressure in the second innings and more over-confident in the first innings. 
It is possible that these results are applicable more to 20 over matches than test matches that have a higher number of dot balls. However, the presence of test match study in our data only biases us away from finding results. High statistical significant indicates an even stronger case.
It is possible that a dot ball and wicket both indicate the ability of the bowler rather than pressure on the batter. If that were the only interpretation, we would have not had any differences in the innings. The fact that there is a difference between dots preceding the wicket in the first innings is not the same as those in the second innings indicates pressure. 
Certain matches are one sided in terms of performance. In those, there would be fewer cases with a dot ball preceding a wicket. This also biases the study against finding results. 
This study can be replicated using an even greater number of deliveries. This study can be sliced along playing teams, gender, location and level of competition.





In a series of papers (@singh2021make ) , I try to examine the following questions.


Some of my other work pertains to perceptions of founders in startups ([@singh2021perception1; @singh2021perception2; @singh2021perception3; @singh2021perception4; @singh2021perception5; @singh2021start]), Olympic medals ([@singh2021olympic; @singh2021names]), Covid data [@singh2020close; @singh2020quantifying], trading strategy [@singh2015square], CSR [@singh2016whether], Crowdfunding [@singh2021crowdfunding; @singh2021emotional] and some Finance and Director diligence stuff [@singh2016impact; @singh2016executive; @singh2017essay] .












